A hierarchical approach for modelling X-ray beamlines. Application to a coherent beamline.

### 2019-06-11 paper-hierarchical.tex
updated Ray's corrections

\begin{abstract}
We consider different approaches to simulate a modern X-ray beamline. Several methodologies with increasing complexity are applied to discuss the relevant parameters that quantify the beamline {\color{green}\sout{performances}performance}. Parameters such as flux, dimensions and intensity distribution of the focused beam and coherence properties are obtained from simple analytical calculations to sophisticated computer simulations using ray tracing and {\color{red}\sout{waveoptics}wave optics} techniques. A {\color{green}\sout{last}latest} generation x-ray nanofocusing beamline for coherent  applications (ID16A at the ESRF) has been chosen  to study in detail the issues related to highly demagnifying  {\color{green}\sout{sychrotron}synchrotron} sources and exploiting the beam coherence. The performance of the beamline is studied for two storage{\color{green}\sout{s}} rings: the old ESRF (emittance 4000 pm) and the new ESRF-EBS (emittance 150 pm). In addition to traditional results in term of flux and beam sizes, an innovative study on the partial coherence properties based on propagation of coherent modes is presented. The different algorithms and methodologies are implemented in the software suite OASYS \cite{codeOASYS}. They are discussed with emphasis beamline {\color{green}\sout{in}placed upon the } benefits and limitations of each. 
\end{abstract}

@@

Many storage ring based x-ray synchrotron facilities are building or planning upgrades to increase brilliance and coherent flux by one to three orders of magnitude.  The first upgrade of a large facility will be the EBS (Extremely Brilliant Source) \cite{orangebook} at the European Synchrotron Radiation Facility (ESRF). The new storage ring of 150 pm emittance that will be commissioned in 2020 will significantly boost the  {\color{green}\sout{x-ray}} coherence of the  {\color{green}\sout{synchrotron} x-ray} beams. Applications exploiting beam coherence such as x-ray photon correlation spectroscopy, coherent diffraction imaging, propagation-based phase contrast imaging and ptychography will strongly benefit from this update.

Accurate calculation and quantitative evaluation of the parameters related to x-ray optics is of {\color{green}\sout{paramount} great} importance for designing, building and exploiting {\color{green}\sout{the}} new beamlines. Every modern beamline follows a procedure of conceptual design plus technical design where detailed simulations are essential. These calculations  {\color{green}\sout{permit to test} allowing testing of} the design parameters in a virtual computer environment where advantages and limitations can be studied quantitatively. {\color{green}\sout{The selection to best performant optics configuration} Selection of the optimal optical layout}  requires the detailed simulation of the optics imperfections, that in many cases are the limiting factor of the beamline optical configuration. {\color{green}\sout{Deformation in} Deformations of} the optical elements due to heat load need to be controlled, therefore optics simulations have to include results from engineering modeling of the thermal deformations {\color{green}\sout{associated to mechanical issues}} usually {\color{red}\sout{dome}done} using finite {\color{red}\sout{e}}element analysis. Traditionally the optics calculations for synchrotron beamlines are done using ray tracing techniques. They model the x-ray beam as a collection of {\color{green}\sout{different}incoherent} rays. {\color{green}\sout{that are incoherent one to another}}  This technique is still essential for obtaining information on aberrations, flux propagation, monochromator bandwidth, etc. However, the high coherence of the new sources {\color{green}\sout{make necessary}can introduce the necessity} to complement the ray tracing methods with models including effects of diffraction and scattering from a coherent beam. A fully coherent {\color{green} hard x-ray}beam is still a dream for the new generation storage rings. In spite of a {\color{green}\sout{high increate of }considerable increase in}  the coherent fraction of the new beams, the coherent fraction at {\color{red}typically\sout{tipically}} hard x-ray photon energies is still a few percent for the new sources.  {\color{green}\sout{This makes still important a } Consequently} ray tracing analysis {\color{green}\sout{that}} does not become obsolete, but has to be complemented with an analysis based on wave optics. Because of the {\color{green}\sout{non perfect} partial} coherence the wave optics analysis becomes non trivial and can be done with different levels of approximation, as discussed in this work. In addition, wave optics simulations are usually computationally expensive.\todo{I noticed that there is very few citations to references in the introduction. This is not so nice, as we don't knowledge any previous work. Maybe one could add a few citations in the introduction.}

To illustrate the applicability of a variety of methods and solutions for beamline optical simulations we have chosen the ESRF beamline ID16A~{\color{red}\cite{ID16A}}. This beamline provides a high-brilliance beam focused to a few nanometers. It displays characteristics increasingly aimed for in modern beamlines: extremely high demagnification (sub-micrometer focused beam size) and high coherence. It combines coherent imaging techniques and X-ray fluorescence microscopy to perform quantitative 3D characterization of the morphology and the elemental composition of specimens in their native state~{\color{red}\cite{XRFID16A}}.

X-ray beamlines are very particular optical systems. Some implications of exploiting the {\color{red}\sout{low}short} wavelengths such as the use of grazing incidence reflectors and the use of perfect crystals as typical hard x-ray monochromators render common commercial software packages ill-adapted for the study of synchrotron beamlines. For this reason many optics tools have been developed by the synchrotron community. Some of them have a long history like SHADOW (more than 40 years old) or SRW (almost 30), to mention only codes used in this work. These historical tools together with other new applications have been collected together in a user friendly software package: OASYS. Many of the calculations presented here have been performed directly in OASYS or with scripts created with the help of OASYS.

The paper is structured as follows: we first introduce in Section~\ref{Beamline description} the main parameters of the beamline and explain the particular choice of the parameters used in the calculations. Section~\ref{Simulations} is the central part of the paper, and describes the different methods used to analyse the beamline with increasing complexity, setting a hierarchy of methodologies that can be exported to the simulations of any beamline. In Section \ref{level0} we calculate with simple methods approximate values of focal size, flux and coherent fraction. Accuracy can improved using geometrical methods in Section \ref{level1} that allow a more complete simulation of the beamline optics within certain limits which are related, in this example, to high demagnification and high coherence of the studied beamline. The wave optics section~\ref{level2} includes different models of increasing complexity as the required accuracy of the results becomes more demanding. \inblue{In the Discussion (Section \ref{discussion}) the results of the different simulation methods are commented. The hierarchical approach allows more accurate and better results but the price to pay is to deal with more complex and computationally expensive tools. Finally Section \ref{summary} summarizes the results.}

@@

This section presents the reader with the main operational parameters for the ID16A beamline at ESRF in two scenarios. This beamline was built during the Upgrade Phase I at ESRF in \inred{2011??}. It uses one or several undulators as source. During the period 2011-2018 it operated exploiting the ESRF storage ring’s double-bend achromat (DBA) lattice with horizontal emittance about 4000 pm. In the second phase of the ESRF Upgrade the original storage ring is being replaced by a new hybrid multi-bend achromat (HMBA) lattice. This new lattice will drastically improve the performances of the ESRF source. The resulting Extreme Brilliant Source (EBS) \cite{orangebook} will increase the brightness and coherent fraction by a factor of more than 100, by strongly reducing the horizontal emittance. This is possible using almost the same ring circumference and electron energy $E_e$. As a result, the x-ray sources will be much coherent, specially in the horizontal direction. In this paper, some quantitative results on how the storage ring upgrade will affect a particular beamline are presented. 
%RC190404: @PhaseI -> from the orange book: On the basis of the Purple Book, the ESRF Council decided in June 2008 to launch the ESRF UP by establishing a two-phase process. The Upgrade Programme Phase I (UP PI) for the period 2009-2015, and the Upgrade Programme Phase II (UP PII) to be prepared in the period 2012-2014, and to be presented to the Council for consideration in 2014 and launch in 2015.
% The  Upgrade  Programme  Phase  I  has  been  on  the  ESFRI  Roadmap  since  its  fi rst publication  in  2006,  and  is  today  heading  towards  its  successful  completion  by  the end of 2015.

ID16A \cite{ID16A} is a beamline that provides a nanometric, highly coherent beam with broad energy-bandpass. It is dedicated to nanofocusing applications. \todo{please summarize some use and application}. The beamline can operate at two fixed photon energies: 17050 eV and 33600 eV. The straight section in the magnet lattice houses four undulators, two 18 mm and another two of 22.4 mm period. The undulator parameters and $K$-values at the resonances for the energies in use are shown in Table \ref{tab:Undulators}.

@@

\begin{table}\label{tab:Undulators}
\centering
\caption{Parameters of the undulators available at the ID16A beamline: period ($\lambda_u$), length ($L$), number of identical undulators ($N$) and deflection parameter ($K$) values for the three main photon energies in use. The valuse are shown for the two ESRF magnetic lattices: the old one (labelled ESRF, with $E_e = 6.04$ GeV) and the new one (EBS, with $E_e = 6.00$ GeV). $n$ is the emission harmonic in use for each particular photon energy.}

@@

Simulations were done using the two different storage ring lattices mentioned before: the (already dismantled) ESRF DBA where the undulators are in a high-$\beta$ straight section (larger horizontal emittance) and the new EBS (lower horizontal emittance) straight sections. The parameters for the current and the new lattice are shown in Tables \ref{tab:TableSources} and \ref{tab:eBeam}.

\begin{table}\label{tab:TableSources}
    \caption{ESRF and ESRF-EBS main storage ring parameters. Main parameters of the present DBA lattice (ESRF) and future HMBA lattice (EBS) storage rings \cite{orangebook}.}
        \begin{tabular}{rcc}
                                              & \textbf{ESRF} & \textbf{EBS}      \\ 
        Lattice type:                         & DBA           & HMBA              \\
        Circunference {[}m{]}:                & 844.930       & 843.979           \\
        Beam energy {[}GeV{]}:                & 6.04          & 6.00              \\
        Beam current {[}mA{]}:                & 200           & 200               \\
        Natural emittance {[}pm$\cdot$rad{]}: & 4000          & 147               \\
        Energy spread {[}$\%${]}:             & 0.0011        & 0.00093           \\       
    \end{tabular}
\end{table}

\begin{table}\label{tab:eBeam}
    \centering
    \caption{ESRF and EBS main electron beam parameters. Electron beam parameters for the high-$\beta$ (ESRF) and the EBS straight sections. Values taken at the symmetry point of the straight section, where the insertion devices are placed. Values for ESRF are taken from \cite{ESRF2014}. For EBS, two (very close) design values are used, from Ref.~\cite{ESRF2014} labelled EBS-S28A, and those coming from further lattice refinement, labelled EBS-S28D. \todo{Remove ESRF-AT and EBS-S28A} }
        \resizebox{\textwidth}{!}{%
        \begin{tabular}{ccccc}
         Lattice & $\sigma_e$ horizontal {[}$\mu$m{]} &  $\sigma_e'$ horizontal {[}$\mu$rad{]} & $\sigma_e$ vertical {[}$\mu$m{]} & $\sigma_e'$ vertical {[}$\mu$rad{]} \\
         \input{tableElectron.txt}

@@

For all simulations presented hereafter, we selected the configuration using a single undulator U18.3 placed in the center of the straight section and tuned to its first harmonic at a photon energy of $E=17225$ eV (\inblue{this value is not identical but close to the usual operation value shown in }Table \ref{tab:Undulators}). This section is divided in three subsections, that present three main levels of simulations: simple calculations that can be made by hand (section \ref{level0}), ray-tracing simulations (section \ref{level1}) and wave optics simulations (section \ref{level2}).

@@

The source size (or divergence) is usually approximated supposing a Gaussian distribution of the electron {\color{green}\sout{sizes}positions} (or divergence) (Table \ref{tab:eBeam}) and {\color{green}\sout{undulator natural}single electron photon} source size or divergence. In some cases, such as for undulator emission, this Gaussian approximation may not be accurate and have some implications that will be discussed later. Let $\sigma_{e(h,v)}$ and $\sigma_{e(h,v)}'$ be the standard deviations of electron beam size and divergence  at the centre of  symmetry of the straight section in a plane perpendicular to the main propagation direction. The subscripts $h$ and $v$ stand for horizontal and vertical, respectively.

@@

Using the values for U18.3, $n$=1, and  $E=17225$ eV (Table \ref{tab:Undulators}) and the optical configuration described in Table \ref{tab:Positions}, one obtains a focal size 139$\times$5 nm$^2$ (H$\times$V) for ESRF and 10$\times$5 nm$^2$ for EBS (Table~\ref{tab:HandCalculations}). The horizontal focal spot for the ESRF lattice is larger than the specified target value for the beamline, therefore the VSS could be closed to 50$\mu$m thus reducing the beamsize at the VSS plane by a factor ca. 8 \inred{$\approx$ 403.8/50=8.076}. Therefore, for ESRF, the horizontal size at the sample will be affected by the same factor obtaining roughly 17$\times$5 nm$^2$

@@

\begin{table}\label{tab:HandCalculations}
    \centering
    \caption{Photon beam size and divergence for selected ID16A beamline positions. Values below were obtained considering undulator U18.3 tuned to its first harmonic at the photon energy of $E=17225$ eV for both the high-$\beta$ (ESRF) and the EBS straight sections. Values are FWHM. Values in parentheses correspond to closing the VSS slit to 50$\mu$m (21.3 if approximating by the Gaussian $\sigma$).}
    \resizebox{\textwidth}{!}{%
    \begin{tabular}{r|ccc|ccc}
                            &               & \textbf{ESRF (high-$\beta$)} &   &         & \textbf{EBS}  &    \\ 
    Element:                & Undulator     & VSS          & Sample            & Undulator & VVS   & Sample       \\
    Position [m]:           & 0.00          & 40.00        & 185.00            & 0.00    & 40.00   & 185.00       \\
    $FWHM_h$ [$\mu$m]:      & 977.19        & 404.00 (50.0)& 0.139 (0.017)     & 71.26   & 29.46   & 0.0102       \\
    $FWHM_v$ [$\mu$m]:      & 9.60          & 482          & 0.0052            & 10.02   & 486.5   & 0.0054       \\
    \newline
    $FWHM'_h$ [$\mu$rad]:   & 26.96         & \inred{65.22?}        & 189062   & 15.60   & 37.74   & 109404       \\
    $FWHM'_v$ [$\mu$rad]:   & 12.04         & 12.04                 & 22264    & 12.16   & 12.16   & 22489        \\                                                   
    \end{tabular}%
    }
\end{table}

Next, we want to estimate the number of photons at the sample. For that, we need the spectrum at the source. It is not straightforward to get this information analytically but one can calculate the flux and power using different codes. Perhaps the most advanced and well maintained codes used in the synchrotron community are SPECTRA \cite{codeSPECTRA} and SRW \cite{codeSRW}. The XOPPY package in the OASYS suite implements three codes: US \cite{codeUS}, URGENT \cite{codeURGENT} and SRW. Figure~\ref{fig:FluxU18} shows the spectra calculated \todo{indicate which code used} over an aperture of 1 $\times$ 1 mm$^2$ placed at $d$=27.1 m. This aperture opening is set to fully accept the central cone at the resonance peak of the harmonic in use ($\max(\Sigma') \times d = 0.73$ mm), with $\Sigma'$ values from in Table~\ref{tab:HandCalculations}. The peak intensity is about  $2.9\cdot10^{14}$ photons/s/0.1{\%}bw for EBS and 1.9 10$^{14}$ for ESRF (see \ref{fig:FluxU18}). One can appreciate how the emittance plays an important role in both peak width and specially in peak value. The number of photons per second after the multilayer monochromator is 2.9 10$^{15}$ for EBS and 1.9 10$^{15}$ for ESRF, considering that the energy bandwidth transmitted by a multilayer monochromator is $\Delta E/E \approx 10^{-2}$ and the bandwidth in Fig.~\ref{fig:FluxU18} is 10$^{-3}$, and supposing, at this point, ideal reflectivity (100\%) for the multilayers.

@@

To estimate the geometrical {\color{red}\sout{transmitivity}transmission?} of the beamline we have to consider which optical elements crop the beam and remove photons. This can be analysed using their angular acceptance or numerical aperture. The multilayer monochromator accepts the whole beam. We consider a VSS closed to $S_h=50~\mu$m in horizontal and fully opened in vertical. For ESRF its {\color{red}\sout{transmittivity}transmission?} is 1/8 (as discussed before) and 1 for EBS. In a similar way, the KB mirrors will crop the beam to a size that is the projection of their {\color{red}useful optical} length onto a direction perpendicular to the optical axis. Using a mirror length of $L_v$=60 mm and $L_h$=26 mm for VFM and HFM, respectively, and incident angle of 15 mrad, the KB behaves as an aperture of $900 \times 390 \mu m^2$. We get N.A. values of 6.2$\times$21.6 $\mu rad^2$ (aperture size divided by the distance from the mirror to the VSS in H and to the source in V). The {\color{red}\sout{transmittivity}transmission?} coefficient is the ratio of the accepted N.A. over the beam divergence, resulting 9.5\% (H) and 0.18\% (V)for ESRF and 16.3\% and 0.18\% for EBS. The overall transmission is the product of H and V ratios, therefore 0.2\% for ESRF and 2.95\% for EBS, therefore 3.8 $10^{12}$ photons/s for ESRF and 8.5 $10^{13}$ photons/s for EBS. 
\todo{check with references $10^9$ at 33.6 keV. I get: 1.6e+15 * 0.00295 * 0.7**3 = 1.62e+12. How closed are the slits? We used the undulator peak value over the whole ML bandwidth, that overestimates the integrated flux.}
\todo{need to consider ML reflectivity for the 3 reflections, also undulator is not perfect, horizontal slit may be closed to less than $50 \mu m$, also KB mirrors may not be fully illuminated and slope error losses due to initial ML mono are neglected}

>>>>

\begin{abstract}
We consider different approaches to simulate a modern X-ray beamline. Several methodologies with increasing complexity are applied to discuss the relevant parameters that quantify the beamline performance. Parameters such as flux, dimensions and intensity distribution of the focused beam and coherence properties are obtained from simple analytical calculations to sophisticated computer simulations using ray tracing and wave optics techniques. A latest generation x-ray nanofocusing beamline for coherent  applications (ID16A at the ESRF) has been chosen  to study in detail the issues related to highly demagnifying  synchrotron sources and exploiting the beam coherence. The performance of the beamline is studied for two storage rings: the old ESRF-1 (emittance 4000 pm) and the new ESRF-EBS (emittance 150 pm). In addition to traditional results in term of flux and beam sizes, an innovative study on the partial coherence properties based on propagation of coherent modes is presented. The different algorithms and methodologies are implemented in the software suite OASYS \cite{codeOASYS}. They are discussed with emphasis beamline placed upon the benefits and limitations of each. 
\end{abstract}

@@

Many storage ring based x-ray synchrotron facilities are building or planning upgrades to increase brilliance and coherent flux by one to three orders of magnitude.  The first upgrade of a large facility will be the EBS (Extremely Brilliant Source) \cite{orangebook} at the European Synchrotron Radiation Facility (ESRF). The new storage ring of 150 pm emittance that will be commissioned in 2020 will significantly boost the coherence of the x-ray beams. Applications exploiting beam coherence such as x-ray photon correlation spectroscopy, coherent diffraction imaging, propagation-based phase contrast imaging and ptychography will strongly benefit from this update.

Accurate calculation and quantitative evaluation of the parameters related to x-ray optics is of great importance for designing, building and exploiting new beamlines. Every modern beamline follows a procedure of conceptual design plus technical design where detailed simulations are essential. These calculations  allowing testing of the design parameters in a virtual computer environment where advantages and limitations can be studied quantitatively. Selection of the optimal optical layout requires the detailed simulation of the optics imperfections, that in many cases are the limiting factor of the beamline optical configuration. Deformations of the optical elements due to heat load need to be controlled, therefore optics simulations have to include results from engineering modeling of the thermal deformations usually done using finite element analysis. Traditionally the optics calculations for synchrotron beamlines are done using ray tracing techniques. They model the x-ray beam as a collection of incoherent rays. This technique is still essential for obtaining information on aberrations, flux propagation, monochromator bandwidth, etc. However, the high coherence of the new sources can introduce the necessity to complement the ray tracing methods with models including effects of diffraction and scattering from a coherent beam. A fully coherent hard x-ray beam is still a dream for the new generation storage rings. In spite of a considerable increase in the coherent fraction of the new beams, the coherent fraction at typically hard x-ray photon energies is still a few percent for the new sources. Consequently ray tracing analysis does not become obsolete, but has to be complemented with an analysis based on wave optics. Because of the partial coherence the wave optics analysis becomes non trivial and can be done with different levels of approximation, as discussed in this work. In addition, wave optics simulations are usually computationally expensive.

To illustrate the applicability of a variety of methods and solutions for beamline optical simulations we have chosen the ESRF beamline ID16A~\cite{ID16A}. This beamline provides a high-brilliance beam focused to a few nanometers. It displays characteristics increasingly aimed for in modern beamlines: extremely high demagnification (sub-micrometer focused beam size) and high coherence. It combines coherent imaging techniques and X-ray fluorescence microscopy to perform quantitative 3D characterization of the morphology and the elemental composition of specimens in their native state~\ingreen{\cite{XRFID16A}}.

X-ray beamlines are very particular optical systems. Some implications of exploiting the short wavelengths such as the use of grazing incidence reflectors and the use of perfect crystals as typical hard x-ray monochromators render common commercial software packages ill-adapted for the study of synchrotron beamlines. For this reason many optics tools have been developed by the synchrotron community. Some of them have a long history like SHADOW (more than 40 years old) or SRW (almost 30), to mention only codes used in this work. These historical tools together with other new applications have been collected together in a user friendly software package: OASYS. Many of the calculations presented here have been performed directly in OASYS or with scripts created with the help of OASYS.

The paper is structured as follows: we first introduce in Section~\ref{Beamline description} the main parameters of the beamline and explain the particular choice of the parameters used in the calculations. Section~\ref{Simulations} is the central part of the paper, and describes the different methods used to analyse the beamline with increasing complexity, setting a hierarchy of methodologies that can be exported to the simulations of any beamline. In Section \ref{level0} we calculate with simple methods approximate values of focal size, flux and coherent fraction. Accuracy can improved using geometrical methods in Section \ref{level1} that allow a more complete simulation of the beamline optics within certain limits which are related, in this example, to high demagnification and high coherence of the studied beamline. The wave optics section~\ref{level2} includes different models of increasing complexity as the required accuracy of the results becomes more demanding. In the Discussion (Section \ref{discussion}) the results of the different simulation methods are commented. The hierarchical approach allows more accurate and better results but the price to pay is to deal with more complex and computationally expensive tools. Finally Section \ref{summary} summarizes the results.

@@

This section presents the reader with the main operational parameters for the ID16A beamline at ESRF in two scenarios. This beamline was built during the Upgrade Phase I at ESRF (2009-2015). It uses one or several undulators as source. During the period 2011-2018 it operated exploiting the ESRF-1 storage ring’s double-bend achromat (DBA) lattice with horizontal emittance about 4000 pm. In the second phase of the ESRF Upgrade the original storage ring is being replaced by a new hybrid multi-bend achromat (HMBA) lattice. This new lattice will drastically improve the performances of the ESRF source. The resulting Extreme Brilliant Source (EBS) \cite{orangebook} will increase the brightness and coherent fraction by a factor of more than 100, by strongly reducing the horizontal emittance. This is possible using almost the same ring circumference and electron energy $E_e$. As a result, the x-ray sources will be much coherent, specially in the horizontal direction. In this paper, some quantitative results on how the storage ring upgrade will affect a particular beamline are presented. 
%RC190404: @PhaseI -> from the orange book: On the basis of the Purple Book, the ESRF Council decided in June 2008 to launch the ESRF UP by establishing a two-phase process. The Upgrade Programme Phase I (UP PI) for the period 2009-2015, and the Upgrade Programme Phase II (UP PII) to be prepared in the period 2012-2014, and to be presented to the Council for consideration in 2014 and launch in 2015.
% The  Upgrade  Programme  Phase  I  has  been  on  the  ESFRI  Roadmap  since  its  fi rst publication  in  2006,  and  is  today  heading  towards  its  successful  completion  by  the end of 2015.

ID16A \cite{ID16A} is a beamline that provides a nanometric, highly coherent beam with broad energy-bandpass. It is dedicated to nanofocusing applications. The beamline can operate at two fixed photon energies: 17050 eV and 33600 eV. The straight section in the magnet lattice houses four undulators, two 18 mm and another two of 22.4 mm period. The undulator parameters and $K$-values at the resonances for the energies in use are shown in Table \ref{tab:Undulators}.

@@

\begin{table}\label{tab:Undulators}
\centering
\caption{Parameters of the undulators available at the ID16A beamline: period ($\lambda_u$), length ($L$), number of identical undulators ($N$) and deflection parameter ($K$) values for the three main photon energies in use. The valuse are shown for the two ESRF magnetic lattices: the old ESRF-1 (labelled ESRF, with $E_e = 6.04$ GeV) and the new ESRF-EBS (labelled EBS, with $E_e = 6.00$ GeV). $n$ is the emission harmonic in use for each particular photon energy.}

@@

Simulations were done using the two different storage ring lattices mentioned before: the (already dismantled) ESRF-1 where the undulators are in a high-$\beta$ straight section (larger horizontal emittance) and the new EBS (lower horizontal emittance) straight sections. The parameters for the current and the new lattice are shown in Tables \ref{tab:TableSources} and \ref{tab:eBeam}.

\begin{table}\label{tab:TableSources}
    \caption{ESRF and ESRF-EBS main storage ring parameters. Main parameters of the present DBA lattice (ESRF-1) and future HMBA lattice (EBS) storage rings \cite{orangebook}.}
        \begin{tabular}{rcc}
                                              & \textbf{ESRF-1} & \textbf{EBS}      \\ 
        Lattice type:                         & DBA           & HMBA              \\
        Circunference {[}m{]}:                & 844.930       & 843.979           \\
        Beam energy {[}GeV{]}:                & 6.04          & 6.00              \\
        Beam current {[}mA{]}:                & 200           & 200               \\
        Natural emittance {[}pm$\cdot$rad{]}: & 4000          & 147               \\
        Energy spread {[}$\%${]}:             & 0.0011        & 0.00093           \\       
    \end{tabular}
\end{table}

\begin{table}\label{tab:eBeam}
    \centering
    \caption{ESRF-1 and EBS main electron beam parameters. Electron beam parameters for the high-$\beta$ (labelled ESRF) and the EBS straight sections. Values taken at the symmetry point of the straight section, where the insertion devices are placed. Values for ESRF are taken from \cite{ESRF2014}. For EBS, two (very close) design values are used, from Ref.~\cite{ESRF2014} labelled EBS-S28A, and those coming from further lattice refinement, labelled EBS-S28D. \todo{Remove ESRF-AT and EBS-S28A} }
        \resizebox{\textwidth}{!}{%
        \begin{tabular}{ccccc}
         Lattice & $\sigma_e$ horizontal {[}$\mu$m{]} &  $\sigma_e'$ horizontal {[}$\mu$rad{]} & $\sigma_e$ vertical {[}$\mu$m{]} & $\sigma_e'$ vertical {[}$\mu$rad{]} \\
         \input{tableElectron.txt}

@@

For all simulations presented hereafter, we selected the configuration using a single undulator U18.3 placed in the center of the straight section and tuned to its first harmonic at a photon energy of $E=17225$ eV (a value close to the usual operation value shown in Table \ref{tab:Undulators}). This section is divided in three subsections, that present three main levels of simulations: simple calculations that can be made by hand (section \ref{level0}), ray-tracing simulations (section \ref{level1}) and wave optics simulations (section \ref{level2}).

@@

The source size (or divergence) is usually approximated supposing a Gaussian distribution of the electron positions (or divergence) (Table \ref{tab:eBeam}) and single-electron photon source size or divergence. In some cases, such as for undulator emission, this Gaussian approximation may not be accurate and have some implications that will be discussed later. Let $\sigma_{e(h,v)}$ and $\sigma_{e(h,v)}'$ be the standard deviations of electron beam size and divergence  at the centre of  symmetry of the straight section in a plane perpendicular to the main propagation direction. The subscripts $h$ and $v$ stand for horizontal and vertical, respectively.

@@

Using the values for U18.3, $n$=1, and  $E=17225$ eV (Table \ref{tab:Undulators}) and the optical configuration described in Table \ref{tab:Positions}, one obtains a focal size 139$\times$5 nm$^2$ (H$\times$V) for ESRF-1 and 10$\times$5 nm$^2$ for EBS (Table~\ref{tab:HandCalculations}). The horizontal focal spot for the ESRF-1 lattice is larger than the specified target value for the beamline, therefore the VSS could be closed to 50$\mu$m thus reducing the beamsize at the VSS plane by a factor ca. 8.
%\inred{8 $\approx$ 403.8/50=8.076}
Therefore, for ESRF-1, the horizontal size at the sample will be affected by the same factor obtaining roughly 17$\times$5 nm$^2$

@@

\begin{table}\label{tab:HandCalculations}
    \centering
    \caption{Photon beam size and divergence for selected ID16A beamline positions. Values below were obtained considering undulator U18.3 tuned to its first harmonic at the photon energy of $E=17225$ eV for both the high-$\beta$ (ESRF) and the EBS straight sections. Values are FWHM. Values in parentheses correspond to closing the VSS slit to 50$\mu$m (21.3 if approximating by the Gaussian $\sigma$).}
    \resizebox{\textwidth}{!}{%
    \begin{tabular}{r|ccc|ccc}
                            &               & \textbf{ESRF (high-$\beta$)} &   &         & \textbf{EBS}  &    \\ 
    Element:                & Undulator     & VSS          & Sample            & Undulator & VVS   & Sample       \\
    Position [m]:           & 0.00          & 40.00        & 185.00            & 0.00    & 40.00   & 185.00       \\
    $FWHM_h$ [$\mu$m]:      & 977.19        & 404.00 (50.0)& 0.139 (0.017)     & 71.26   & 29.46   & 0.0102       \\
    $FWHM_v$ [$\mu$m]:      & 9.60          & 482          & 0.0052            & 10.02   & 486.5   & 0.0054       \\
    \newline
    $FWHM'_h$ [$\mu$rad]:   & 26.96         & 65.22        & 189062   & 15.60   & 37.74   & 109404       \\
    $FWHM'_v$ [$\mu$rad]:   & 12.04         & 12.04                 & 22264    & 12.16   & 12.16   & 22489        \\                                                   
    \end{tabular}%
    }
\end{table}

Next, we want to estimate the number of photons at the sample. For that, we need the spectrum at the source. It is not straightforward to get this information analytically but one can calculate the flux and power using different codes. Perhaps the most advanced and well maintained codes used in the synchrotron community are SPECTRA \cite{codeSPECTRA} and SRW \cite{codeSRW}. The XOPPY package in the OASYS suite implements three codes: US \cite{codeUS}, URGENT \cite{codeURGENT} and SRW. Figure~\ref{fig:FluxU18} shows the spectra calculated using SRW over an aperture of 1 $\times$ 1 mm$^2$ placed at $d$=27.1 m. This aperture opening is set to fully accept the central cone at the resonance peak of the harmonic in use ($\max(\Sigma') \times d = 0.73$ mm), with $\Sigma'$ values from in Table~\ref{tab:HandCalculations}. The peak intensity is about  $2.9\cdot10^{14}$ photons/s/0.1{\%}bw for EBS and 1.9 10$^{14}$ for ESRF-1 (see \ref{fig:FluxU18}). One can appreciate how the emittance plays an important role in both peak width and specially in peak value. The number of photons per second after the multilayer monochromator is 2.9 10$^{15}$ for EBS and 1.9 10$^{15}$ for ESRF-1, considering that the energy bandwidth transmitted by a multilayer monochromator is $\Delta E/E \approx 10^{-2}$ and the bandwidth in Fig.~\ref{fig:FluxU18} is 10$^{-3}$, and supposing, at this point, ideal reflectivity (100\%) for the multilayers.

@@

To estimate the geometrical transmission of the beamline we have to consider which optical elements crop the beam and remove photons. This can be analysed using their angular acceptance or numerical aperture. The multilayer monochromator accepts the whole beam. We consider a VSS closed to $S_h=50~\mu$m in horizontal and fully opened in vertical. For ESRF its transmission is 1/8 (as discussed before) and 1 for EBS. In a similar way, the KB mirrors will crop the beam to a size that is the projection of their useful optical length onto a direction perpendicular to the optical axis. Using a mirror length of $L_v$=60 mm and $L_h$=26 mm for VFM and HFM, respectively, and incident angle of 15 mrad, the KB behaves as an aperture of $900 \times 390 \mu m^2$. We get N.A. values of 6.2$\times$21.6 $\mu rad^2$ (aperture size divided by the distance from the mirror to the VSS in H and to the source in V). The transmission coefficient is the ratio of the accepted N.A. over the beam divergence, resulting 9.5\% (H) and 0.18\% (V)for ESRF and 16.3\% and 0.18\% for EBS. The overall transmission is the product of H and V ratios, therefore 0.2\% for ESRF and 2.95\% for EBS, therefore 3.8 $10^{12}$ photons/s for ESRF and 8.5 $10^{13}$ photons/s for EBS. 
%\todo{check with references $10^9$ at 33.6 keV. I get: 1.6e+15 * 0.00295 * 0.7**3 = 1.62e+12. How closed are the slits? We used the undulator peak value over the whole ML bandwidth, that overestimates the integrated flux.}
%\todo{need to consider ML reflectivity for the 3 reflections, also undulator is not perfect, horizontal slit may be closed to less than $50 \mu m$, also KB mirrors may not be fully illuminated and slope error losses due to initial ML mono are neglected}

One can also make some simple inferences about the beam coherence. The coherent fraction (CF) \cite{arxivCF} of the undulator beam is the occupation of the lowest coherent mode, and will be discussed further in Section~\ref{level2}. An approximated estimation of CF can be obtained using Eqs.~\ref{eq:photon small sigmas} and \ref{eq:photon big sigmas}. The coherent fraction is the ratio of the phase space of the fully coherent (zero emittance) beam raletive to the true beam. The CF in the horizontal and vertical planes are
\begin{equation}\label{eq:coherent fraction}
 CF_{(h,v)} = \frac{\sigma_u \sigma'_u}{\Sigma_{(h,v)} \Sigma'_{(h,v)}},
\end{equation}
and the total coherent fraction is the product of both (CF=CF$_h \times $CF$_v$). An estimation of the CF calculated in this way gives CF=0.12\% for the ESRF and 2.73\% for EBS at 17 keV photon energy. The CF roughly indicates the ratio of the ``coherent photons'' over the total number of photons. It is a good indicator of the quality of a coherent beam (a fully coherent beam has CF=1 and an incoherent beam CF=0). The beamline optics act as a ``coherence filter'' by removing the ``bad'' or ``incoherent'' photons and in consequence increasing the CF. A complete cleaning is impossible because of the spatial overlapping of the different coherence modes. However, apertures centered at the optical axis would clean the beam from these undesired photons. If we compare the values of the beamline transmission due to geometrical aperturing  (0.2\% for ESRF-1 and 2.95\% for EBS) with the the CF values, we remark that they are very close. This suggests that the beam at the focal position will be highly coherent because the fraction of the photons removed from the beam matches the CF value that quantify the fraction of ``good'' or ``coherent'' photons.

To assess the focused beam size given by a coherent beam, one can consider the beamline in two sections: First, the ML monochromator and VSS that produce a coherent plane wave as the source (vertical) and VSS (horizontal) are both very far from the entrance aperture of the focusing optics. Second, the KB-system playing a dual role, acting as a focusing element of focal distance $f$=5 cm, but also cropping the beam by acting as a square aperture of $D \times D=$ 400$\times$400 $\mu m^2$ (the length of the mirror projected in a plane perpendicular to the beam propagation).
%\todo{REMOVE: we obtained 900 in H, that will be reduced by 0.5 as propagated from 10 cm to 5 cm upstream of the sample, and 390 in V approximated by 400}.
This aperture has two effects: i) as discussed before, it removes a lot of off-axis photons increasing the CF to a value close to one, and ii) broadens of the focal spot because of coherent diffraction. The diffraction by a square aperture of a collapsing spherical wavefront has an intensity distribution proportional to $\sinc^2(k D x/2 f) \sinc^2(k D y / 2 f)$, \todo{include citation?} where $k=2\pi/\lambda$, and $x$ ($y$) is the horizontal (vertical) coordinate. Considering that the FWHM of $\sinc(x)$ is approximately 2.78, one obtains a FWHM of 7.75$\times$7.75 nm$^2$ for the focal spot. This spot size is of the order or larger than that calculated geometrical considerations alone, and therefore contributes significantly to the overall size of the focused beam. One can conclude that for the conditions in use this is a diffraction limited beamline.
